---
title: "R Introduction"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
turnout <- read.csv("http://qss.princeton.press/student-files/INTRO/turnout.csv")
```


## Loading Data

### Load and inspect data 

*Here's an exercise to familiarize yourself with loading data in R*

We are going to use the U.S. election turnout data set from Kosuke Imai's *Quantitative Social Science* book. This is available at the url: (http://qss.princeton.press/student-files/INTRO/turnout.csv)

For the first exercise, load the data into R. You can do this either by downloading the file to your working directory, setting the working directory, and opening the file OR you can supply the read function directly with the url, with the url in quotations "http://qss.princeton.press/student-files/INTRO/turnout.csv"). Hint: note that it is a .csv file. Store the data as an object called `turnout`.

Then, report how many rows and how many columns are in the data.

```{r exercise1, exercise = TRUE}

```


```{r exercise1-solution}
turnout <- read.csv("http://qss.princeton.press/student-files/INTRO/turnout.csv")
nrow(turnout)
ncol(turnout)
```

*Let's inspect the data*

Show the first 6 rows of the `turnout` dataframe. Note how this is a rectangular dataset, where each column is a different variable, and each row corresponds to a different year's set of data.

```{r previewdata, exercise=TRUE, exercise.lines = 1}

```

```{r previewdata-hint}
head()
```


```{r previewdata-solution}
head(turnout)
```


We can look at the names of the variables using the `names()` function in R. Each variable can be accessed by writing the name of the dataframe followed by a \$. Find the year variable. What year is in the second entry of this variable?

```{r brackets, exercise = TRUE, exercise.lines = 4}
names(turnout)

```

```{r brackets-solution}
names(turnout)
turnout$year[2]
```

## Summarizing Data


### Taking the mean and standard deviation

*Here's an exercise to perform basic summary functions.*

Take the mean of the variable that represents the voting eligible population `VEP`. Then, take its standard deviation.

```{r meanvep, exercise=TRUE, exercise.lines = 4}

```

```{r meanvep-solution}
mean(turnout$VEP)
sd(turnout$VEP)
```

## Manipulating Variables

*Here's an exercise to start to combine and manipulate variables*

Subtract felons and noncitizens from the voting age population. Store this as a new variable called `turnout$adjustedVAP`. Hint: you will need to use the assignment tool `<-` to store the variable. Compare the mean of the adjusted and unadjusted VAP.

```{r newvar, exercise=TRUE, exercise.lines = 4}

```

```{r newvar-solution}
turnout$adjustedVAP <- turnout$VAP - turnout$felons - turnout$noncit

mean(turnout$VAP)
mean(turnout$adjustedVAP)
```

Calculate the VEP turnout rate for each year by dividing the `total` turnout by VEP. What is the mean VEP turnout rate, as a percent?

```{r division, exercise=TRUE, exercise.lines=5}

```

```{r division-solution}
turnout$VEPrate <- turnout$total / turnout$VEP
mean(turnout$VEPrate)*100
```

The `ANES` variable includes the turnout rate reported in a national survey each year. This is based on whether people self-report they voted, while the VEP rate is based on actual data. Compare the average turnout on the ANES with the average VEP rate you just calculated. Note: each exercise is independent, so if you want to use code from the previous exercise, you will need to paste it below.

```{r anescompare, exercise=TRUE, exercise.lines=5}

```


```{r anescompare-solution}
turnout$VEPrate <- turnout$total / turnout$VEP
mean(turnout$VEPrate)*100

mean(turnout$ANES)
```



```{r quiz}
quiz(
  question("What might be one reasonable interpretation?",
    answer("People underreport their turnout on national surveys."),
    answer("The rates are equal, suggesting the ANES and VEP rates reflect the same calculation"),
    answer("People tend to overreport their turnout on national surveys.", correct = TRUE)
  ))
```
